---
title: "Introduction to spconform"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to spconform}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 6,
  fig.height = 5,
  fig.align = "center",
  eval = requireNamespace("sp", quietly = TRUE)
)
```

```{r setup}
library(spconform)
```

## Overview

`spconform` provides distribution-free, finite-sample prediction
intervals for spatial and spatio-temporal data by relaxing the
exchangeability assumption of standard conformal prediction. It offers
two core procedures:

- `scp_geostatistical()` for point-referenced (geostatistical) data,
  using locally weighted split conformal prediction with spatial (and,
  optionally, temporal) Gaussian kernels.
- `scp_areal()` for areal (lattice) data, using a neighbourhood-weighted
  leave-one-out conformal procedure based on a graph adjacency
  structure.

Both procedures are **model-agnostic**: you supply your own point
predictor (a linear model, a GAM, a random forest, kriging, or anything
else), and `spconform` handles only the conformal calibration layer,
producing prediction intervals with a guaranteed marginal coverage
level regardless of whether your predictor is well specified.

This vignette illustrates both procedures on the `meuse` river dataset
(Pebesma and Bivand 2005), a standard geostatistics benchmark
distributed with the **sp** package.

```{r}
library(sp)
data(meuse)

s <- as.matrix(meuse[, c("x", "y")])
y <- log(meuse$zinc)
```

We use log-transformed zinc concentration (ppm) as the response, and
the projected coordinates as the spatial index. The 155 sampling
locations trace a diagonal band following the natural course of the
river.

```{r fig-layout}
plot(meuse$x, meuse$y, col = rgb(0.2, 0.4, 0.8, 0.5), pch = 19,
     xlab = "X coordinate", ylab = "Y coordinate",
     main = "Meuse Sampling Locations")
```

## Geostatistical (point-referenced) prediction

### Defining a point predictor

`scp_geostatistical()` requires a `pred_fun` with signature
`function(s_train, y_train, s_new)`, returning point predictions at the
new locations `s_new`. Here we use a deliberately simple quadratic
trend surface, fit by ordinary least squares:

```{r}
pred_fun <- function(s_train, y_train, s_new) {
  fit <- lm(y_train ~ s_train[, 1] + s_train[, 2] +
              I(s_train[, 1]^2) + I(s_train[, 2]^2))
  cbind(1, s_new[, 1], s_new[, 2], s_new[, 1]^2, s_new[, 2]^2) %*% coef(fit)
}
```

This predictor is intentionally simple; the point of conformal
prediction is that the resulting intervals remain valid even when the
underlying model is imperfect.

### Fitting the conformal intervals

We split the data 70/30 into training and test sets, and construct 90%
prediction intervals on the test locations:

```{r}
set.seed(1)
n <- nrow(s)
idx <- sample(n, floor(0.7 * n))

s_train <- s[idx, ]; y_train <- y[idx]
s_test  <- s[-idx, ]; y_test  <- y[-idx]

out <- scp_geostatistical(s_train, y_train, s_test, pred_fun,
                           alpha = 0.1, seed = 1)
print(out)
```

`coverage_report()` compares the intervals against the (here, known)
true test values:

```{r}
coverage_report(out, y_test)
```

### Visualizing the intervals

The `plot()` method displays the point predictions, intervals, and
(optionally) the true values:

```{r fig-intervals}
plot(out, y_true = y_test)
```

### Assessing stability via Monte Carlo replication

A single train/test split can be misleading. We repeat the split 50
times to assess whether coverage is stable around the nominal target:

```{r}
set.seed(123)
coverages <- numeric(50)
widths    <- numeric(50)

for (i in 1:50) {
  idx_i <- sample(n, floor(0.7 * n))
  s_tr <- s[idx_i, ]; y_tr <- y[idx_i]
  s_te <- s[-idx_i, ]; y_te <- y[-idx_i]

  out_i <- scp_geostatistical(s_tr, y_tr, s_te, pred_fun,
                               alpha = 0.1, seed = i)

  rep_i <- coverage_report(out_i, y_te)
  coverages[i] <- rep_i$coverage
  widths[i]    <- rep_i$mean_width
}

mean(coverages)
sd(coverages)
mean(widths)
```

```{r fig-coverage-hist}
hist(coverages, breaks = 15, col = "lightblue", border = "white",
     main = "Empirical Coverage Across 50 Random Splits",
     xlab = "Empirical Coverage", xlim = c(0.7, 1))
abline(v = 0.90, col = "red", lwd = 2, lty = 2)
legend("topleft", legend = "Nominal target (0.90)",
       col = "red", lty = 2, bty = "n")
```

The mean coverage across replications is close to the nominal 90%
target, with low variability across data partitions — indicating that
the coverage guarantee is stable and not an artifact of a single split.

### Spatial distribution of interval width

Because `scp_geostatistical()` weights calibration points by proximity
to each target location, interval width can vary spatially, reflecting
local data density and configuration:

```{r fig-spatial-width}
plot_df <- data.frame(
  x = s_test[, 1],
  y = s_test[, 2],
  width = out$upper - out$lower
)

plot(plot_df$x, plot_df$y,
     cex = plot_df$width, pch = 19,
     col = rgb(0.2, 0.4, 0.8, 0.5),
     xlab = "X coordinate", ylab = "Y coordinate",
     main = "Spatial Distribution of Interval Width")
```

## Areal (lattice) prediction

`scp_areal()` targets data observed on a fixed set of areal units (e.g.
counties, grid cells) linked by an adjacency structure, rather than
continuous coordinates. To illustrate this on the same phenomenon, we
aggregate the point-referenced Meuse data onto a regular $6\times6$
grid, retaining occupied cells and taking the mean log-zinc
concentration within each as the areal response.

```{r}
xbreaks <- seq(min(meuse$x), max(meuse$x), length.out = 7)
ybreaks <- seq(min(meuse$y), max(meuse$y), length.out = 7)

meuse$cell_x  <- cut(meuse$x, xbreaks, include.lowest = TRUE, labels = FALSE)
meuse$cell_y  <- cut(meuse$y, ybreaks, include.lowest = TRUE, labels = FALSE)
meuse$cell_id <- (meuse$cell_y - 1) * 6 + meuse$cell_x

agg <- aggregate(log(zinc) ~ cell_id, data = meuse, FUN = mean)
names(agg) <- c("cell_id", "y")

cell_coords <- unique(meuse[, c("cell_id", "cell_x", "cell_y")])
agg <- merge(agg, cell_coords, by = "cell_id")
agg <- agg[order(agg$cell_id), ]

n_cells <- nrow(agg)
adj <- matrix(0, n_cells, n_cells)
for (i in 1:n_cells) {
  for (j in 1:n_cells) {
    if (i != j) {
      dx <- abs(agg$cell_x[i] - agg$cell_x[j])
      dy <- abs(agg$cell_y[i] - agg$cell_y[j])
      if (dx <= 1 && dy <= 1) adj[i, j] <- 1
    }
  }
}
```

`adj` is a binary adjacency matrix linking grid-adjacent cells. We now
apply `scp_areal()` at a nominal 80% coverage level, using the default
neighbourhood-mean predictor:

```{r}
out2 <- scp_areal(agg$y, adjacency = adj, alpha = 0.2)
print(out2)
summary(out2)
coverage_report(out2, agg$y)
```

```{r fig-areal-intervals}
plot(out2, y_true = agg$y)
```

Most areal units show narrow intervals, with occasional exceptions at
units with a sparse neighbourhood (e.g., boundary cells of the grid).
This is a desirable property of `scp_areal()`: units with fewer graph
neighbours have a smaller, less informative local calibration set, and
their wider interval correctly reflects the higher predictive
uncertainty at the periphery of the spatial domain, rather than
understating it.

### Comparing interval widths across procedures

```{r fig-width-comparison}
boxplot(list(Geostatistical = out$upper - out$lower,
             Areal          = out2$upper - out2$lower),
        main = "Interval Width Comparison",
        ylab = "Interval Width",
        col  = c("lightblue", "lightgreen"))
```

## Summary

| Dataset                             | Type           | n   | Target coverage | Empirical coverage |
|---------------------------------------|----------------|-----|-------------------|-----------------------|
| Meuse (zinc, point-referenced)        | Geostatistical | 155 | 0.90              | ~0.90–0.92 (Monte Carlo mean) |
| Meuse (aggregated, 6x6 grid)          | Areal          | 21  | 0.80              | ~0.76–0.81 |

Both procedures achieve empirical coverage close to their nominal
targets on this real environmental dataset, using deliberately simple
underlying predictors (a misspecified trend surface, and a
neighbourhood mean), illustrating that the coverage guarantee comes
from the conformal calibration layer itself rather than from correct
model specification.

## Using your own predictor

Both `scp_geostatistical()` and `scp_areal()` accept an arbitrary
prediction function, so you are not limited to the simple examples
above. For geostatistical data, any function of the form
`function(s_train, y_train, s_new)` returning a numeric vector of
predictions will work — including kriging (e.g., via **gstat**),
generalized additive models (via **mgcv**), or random forests (via
**ranger**). For areal data, you may supply a custom
`function(y_train, X_train, idx_train, idx_target, adjacency)` in place
of the default neighbourhood-mean predictor.

## References

- Mao, H., Martin, R., and Reich, B. J. (2023). Valid Model-Free
  Spatial Prediction. *Journal of the American Statistical Association*,
  118(541), 496–510. doi:10.1080/01621459.2021.2016422
- Pebesma, E. J., and Bivand, R. S. (2005). Classes and Methods for
  Spatial Data in R. *R News*, 5(2), 9-13.
- Vovk, V., Gammerman, A., and Shafer, G. (2005). *Algorithmic Learning
  in a Random World*. Springer.
